Accelerated evolution of convective simulations

### 2018-06-11 tex/bvp_paper.tex
adds new fig 2 file

Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal, periodic) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  
We study 2D and 3D convection in which the domain is a cartesian box, 
whose dimensionless vertical extent is $z \in [-1/2, 1/2]$, 
and which is horizontally periodic with an extent of $x, y \in [0, \Gamma]$,
where $\Gamma = 2$ is the aspect ratio, as has been previously studied
\cite{goluskin&all2014, johnston&doering2009}. 
In 2D simulations, we set $v = \partial_y = 0$.
We specify no-slip, impenetrable boundary conditions at both the top and
bottom boundary,
\begin{equation}
u = v = w = 0 \, \, \text{at}\,\,z = \pm 1/2,
\label{eqn:vel_bcs}
\end{equation}
and we use mixed thermal boundary conditions, such that
\begin{equation}
T_1 = 0 \,\,\text{at}\,\, z=+1/2, \qquad
\frac{\partial T_1}{\partial z} = 0\,\,\text{at}\,\,z=-1/2.
\label{eqn:temp_bcs}
\end{equation}
For this choice of boundary conditions, the critical value of Ra at which
the onset of convection occurs is Ra$_{\text{crit}} = 1295.78$ \cite{goluskin2016}, and the
supercriticality of a run is defined as $S \equiv \text{Ra}/\text{Ra}_{\text{crit}}$.
Studies of convection which aim to model
astrophysical systems such as stars often employ mixed thermal
boundary conditions, as we do here \cite{hurlburt&all1984, cattaneo&all1991, korre&all2017}.
Our choice of the thermal boundary conditions in Eqn.~(\ref{eqn:temp_bcs}) 
was motivated by the fact that accelerated evolution is simpler when both the
thermal profile and the flux through the domain are fixed at a boundary 
(see section \ref{sec:ae}).

The initial temperature profile is linearly unstable,
$T_0(z) = -z$. On top of this profile, we fill $T_1$ with
random white noise whose magnitude is $10^{-6}\mathcal{P}$, and which is
vertically tapered so as to match the thermal boundary conditions.
This ensures that the initial perturbations are much smaller than the
evolved convective temperature perturbations, even at large Ra.
We filter this noise spectrum in coefficient space, 
such that only the lower 25\% of the coefficients
have power; this low-pass filter is used to avoid populating the
highest wavenumbers with noise in order to improve the stability of our
spectral timestepping methods.

@@

\section{Computational Benefits of AE}
\label{sec:speedups}
The small differences between the SE and AE solutions for the case studied in 
Figs. \ref{fig:time_trace}, \ref{fig:temp_comparison}, \& \ref{fig:pdf_comparison},
show the extreme power of AE.  In addition to the fact that AE runs require much
less total simulation time (see e.g., Fig.~\ref{fig:time_trace}a\&c and the time
values in the green highlighted regions), the first application of AE in a given
simulation (Fig.~\ref{fig:time_trace}c, at the arrow labeled ``1'') drastically
increases the average timestep by immediately progressing the simulation into
a more converged state. For the $S = 10^5$ case we examined in detail, the
average time step grew by a factor of 2-3 due to the decreased convective
velocities from the transient state to the evolved state. 
At $S = 10^7$, the AE solve immediately improved the timestep size 
by nearly a factor of 4.
Thus, AE achieves converged solutions in few simulation time units,
while also taking less real world time per simulation time unit, when compared
to the slowly converging SE state at the same simulation time.

>>>>

Variables are time-evolved on a dealiased Chebyshev (vertical)
and Fourier (horizontal, periodic) domain in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  
We study 2D and 3D convection in which the domain is a cartesian box, 
whose dimensionless vertical extent is $z \in [0, 1]$, 
and which is horizontally periodic with an extent of $x, y \in [0, \Gamma]$,
where $\Gamma = 2$ is the aspect ratio, as has been previously studied
\cite{goluskin&all2014, johnston&doering2009}. 
In 2D simulations, we set $v = \partial_y = 0$.
We specify no-slip, impenetrable boundary conditions at both the top and
bottom boundary,
\begin{equation}
u = v = w = 0 \, \, \text{at}\,\,z = \pm 0,1,
\label{eqn:vel_bcs}
\end{equation}
and we use mixed thermal boundary conditions, such that
\begin{equation}
T_1 = 0 \,\,\text{at}\,\, z=1, \qquad
\frac{\partial T_1}{\partial z} = 0\,\,\text{at}\,\,z=0.
\label{eqn:temp_bcs}
\end{equation}
For this choice of boundary conditions, the critical value of Ra at which
the onset of convection occurs is Ra$_{\text{crit}} = 1295.78$ \cite{goluskin2016}, and the
supercriticality of a run is defined as $S \equiv \text{Ra}/\text{Ra}_{\text{crit}}$.
Studies of convection which aim to model
astrophysical systems such as stars often employ mixed thermal
boundary conditions, as we do here \cite{hurlburt&all1984, cattaneo&all1991, korre&all2017}.
Our choice of the thermal boundary conditions in Eqn.~(\ref{eqn:temp_bcs}) 
was motivated by the fact that accelerated evolution is simpler when both the
thermal profile and the flux through the domain are fixed at a boundary 
(see section \ref{sec:ae}).

The initial temperature profile is linearly unstable,
$T_0(z) = 0.5 - z$. On top of this profile, we fill $T_1$ with
random white noise whose magnitude is $10^{-6}\mathcal{P}$, and which is
vertically tapered so as to match the thermal boundary conditions.
This ensures that the initial perturbations are much smaller than the
evolved convective temperature perturbations, even at large Ra.
We filter this noise spectrum in coefficient space, 
such that only the lower 25\% of the coefficients
have power; this low-pass filter is used to avoid populating the
highest wavenumbers with noise in order to improve the stability of our
spectral timestepping methods.

@@

\section{Computational Benefits of AE}
\label{sec:speedups}

\begin{table}
\caption{
}
\label{table:speed}
\begin{center}
\begin{tabularx}{\textwidth}{ X X X X }
\hline																	
$S$	&	nz, nx, ny	&	$t_{\text{sim,AE}}/t_{\text{sim,SE}}$	&	
$t_{\text{CPU,AE}}/t_{\text{CPU,SE}}$		\\
\hline \hline \multicolumn{4}{c}{\vspace{-0.2cm}}\\
\multicolumn{4}{c}{\vspace{0.1cm}2D Runs} \\
\hline
$10^1$	&	32,	64	&	2.32    & \\
$10^2$	&	64,	128	&	0.905   & 2.03  & 4.369 &  2.152    \\
$10^3$	&	128,256	&	0.24    & 0.387 & 20.63 &  53.25    \\
$10^4$	&	256,512	&	0.084   & 0.15  & 180.82&  1205.25  \\
$10^5$	&	512,1024&	0.036   & 0.115 & 2757.73& 24061.75 \\
\\ \hline \hline \multicolumn{4}{c}{\vspace{-0.2cm}}\\
\multicolumn{4}{c}{\vspace{0.1cm}3D Runs} \\
\hline
$10^1$	&	32,	64$\times$64	& 2.29	& 106.15/61.5\\
$10^2$	&	64,	128$\times$128	& 0.69	& 111.92/185.35\\
$10^3$	&	128,256$\times$256	& 0.21	& 1426.39/6985.18\\
$10^4$	&	256,512$\times$512	& 0.068	& 23063.77/331,717\\
\hline																	
\end{tabularx}
\end{center}
\end{table}

The small differences between the SE and AE solutions for the case studied in 
Figs. \ref{fig:time_trace}, \ref{fig:temp_comparison}, \& \ref{fig:pdf_comparison},
show the extreme power of AE.  In addition to the fact that AE runs require much
less total simulation time (see e.g., Fig.~\ref{fig:time_trace}a\&c and the time
values in the green highlighted regions), the first application of AE in a given
simulation (Fig.~\ref{fig:time_trace}c, at the arrow labeled ``1'') drastically
increases the average timestep by immediately progressing the simulation into
a more converged state. For the $S = 10^5$ case we examined in detail, the
average time step grew by a factor of 2-3 due to the decreased convective
velocities from the transient state to the evolved state. 
At $S = 10^7$, the AE solve immediately improved the timestep size 
by nearly a factor of 4.
Thus, AE achieves converged solutions in few simulation time units,
while also taking less real world time per simulation time unit, when compared
to the slowly converging SE state at the same simulation time.
